Constructing a universe for the setoid model

### 2020-06-29 paper.tex
More paper writing

@@

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. One aspect of the
metatheory that we pay attention to is it's minimality, as we want to keep it as
simple as possible. In particular, we want to avoid having to assume strong
definition schemas that go beyond inductive families.

$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via $\textsf{Lift} : \mProp \to \mType$. The universe of propositions is
inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
\mProp$. We also require indexed W-types, both in $\mType$ and $\mProp$:

@@

In addition to the proof-relevant types just listed, we will need the
propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
be defined from their $\mType$ counterparts via truncation. That is, given $P :
\mProp$ and $Q : P \to \mProp$:

@@

Note that a universe of strict propositions has been recently added to the Agda
proof assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact,
most of the developments presented here have been formalized and proof-checked
using Agda extended with strict propositions. For convenience we will employ
Agda notation for metatheoretic constructions throughout the paper, with the
implication that these can be understood as, or easily translated to, equivalent
$MLTT^{Prop}$ definitions.

@@

This gives a complete definition of a universe of setoids. However, as already
pointed out, this definition uses induction-recursion. Moreover, the kind of
recursion involved in this definition is particularly complex, since it involves
several recursive definitions whose types contain mutually defined functions. It
is not clear whether this definition schema, which we could call
induction-recursion-recursion, can be reduced to simpler, better understood
notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
need to extend the metatheory with simple forms of induction-recursion goes
against our objective to keep the metatheory as small and essential as possible.

@@

It is known that finitary inductive-inductive definitions can be translated into
inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough} but it is
not clear whether this extends to an infinitary type like our universe.
%
Of course, the absence of a general reduction method does not mean that
particular concrete instances of induction-induction cannot be reduced to
simpler principles, which is exactly what we hope for our universe construction.

@@

TODO: depends on whether we manage to finish the formalization.

@@

\section{Related work}

>>>>

\subsection{Related work}

@@

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. We employ Agda
notation for metatheoretic terms throughout the paper.

One aspect of the metatheory that we pay attention to is it's minimality, as we
want to keep it as simple as possible. In particular, we want to avoid having to
assume strong definition schemas that go beyond inductive families.

$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via $\textsf{Lift} : \mProp \to \mType$.

%% The universe of propositions is
%% inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
%% \mProp$.

We also require indexed W-types, both in $\mType$ and $\mProp$:

@@

From $W_p$ we can define propositional truncation:

\begin{align*}
  & \|\_\| : \mType \to \mProp \\
  & |\_| : \{A : \mType \} \to A \to \| A \| \\
  & \textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P
\end{align*}

In addition to the above type formers in $\mType$, we will need the
propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
be defined from their $\mType$ counterparts via truncation. That is, given $P :
\mProp$ and $Q : P \to \mProp$:

@@

Finally, we will assume the presence of an identity type in the style of
Martin-L\"of's inductive identity type. The main difference is that our identity
type is a $\mProp$-valued relation. Naturally, we also have a \textsf{transp}
combinator representing the corresponding elimination principle.

\begin{align*}
  & \textsf{Id} : \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} : \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} : \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
\end{align*}

Note that this
represents a strong elimination principle, as it allows to eliminate a strict
proposition into arbitrary types. Indeed, the reason why we must assume this
identity type as primitive is because, even though we could have defined
\textsf{Id} from $W_p$, \textsf{transp} would not be definable from $W_p$
elimination. \todo{what is the justification for this?}

\subsection{Formalization}

A universe of strict propositions has been recently added to the Agda proof
assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact, most of
the developments presented here have been formalized and proof-checked using
Agda extended with strict propositions.
%
For convenience, we also slightly deviate from $MLTT^{Prop}$ both in the paper
and in the formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
instead of their derived forms. We operate under the assumption that everything
can be equivalently carried out in $MLTT^{Prop}$, although we haven't fully
checked all the necessary details.

@@

This gives a complete definition of a universe of setoids. However, as already
pointed out, this definition uses induction-recursion. Moreover, the kind of
recursion involved in this definition is particularly complex, since TODO. It is
not clear whether this definition can be reduced to simpler, better understood
notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
need to extend the metatheory with simple forms of induction-recursion goes
against our objective to keep the metatheory as small and essential as possible.

@@

@@

\subsection{Reducing finitary induction-induction}

It is known that finitary inductive-inductive definitions can be reduced to
inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough}.
%
To illustrate the idea, let us consider a well-known example of finitary
inductive-inductive type, the intrinsic encoding of type theory in type theory
itself. Actually, we only need to consider the type of contexts $\textsf{Con} :
\mType$ and the type of types $\textsf{Ty} : \textsf{Con} \to \mType$; since the
latter is indexed over the former, this is already an example of
induction-induction.

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.

\begin{align*}
  & \bullet : \textsf{Con} \\
  & \_,\_ : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \\
  & \iota : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \Pi : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{align*}

The general method to eliminate induction-induction is to split the original
inductive-inductive types into a type of codes and an associated well-formedness
predicate. In our \textsf{Con/Ty} example, these would be respectively given by
codes $\textsf{Con}_0, \textsf{Ty}_0 : \mType$ and predicates $\textsf{Con}_1 :
\textsf{Con}_0 \to \mType, \textsf{Ty}_1 : \textsf{Con}_0 \to \textsf{Ty}_0 \to
\mType$.

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed.

\begin{align*}
  & \bullet_0 : \textsf{Con}_0 \\
  & \_,_0\_ : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
  & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
  & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{align*}

We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
\Sigma (\GG_0 : \textsf{Con}_0)(\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
:\equiv \Sigma(A_0 :
\textsf{Ty}_0)(\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straighforward:

\begin{align*}
  & \bullet :\equiv (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) :\equiv ((\GG_0 ,_0 A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) :\equiv (\iota_0 \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) :\equiv
    ((\Pi_0\ \GG_0\ A_0\ B_0) , (\Pi_1\ \GG_1\ A_1\ B_1))
\end{align*}

Finally, we can define eliminators/induction principles for $\textsf{Con}$ and
$\textsf{Ty}$ as just defined, by induction on the well-typing predicates.

For any inductive-inductive type $A : \mType, B: A \to \mType$, the most general
form of (dependent) eliminator is given by motives $F : A \to \mType$ and $G :
(a : A) \to (b : B a) \to F\ a \to \mType$, and the functions $\textsf{elim}_A :
(a : A) \to F\ a, \textsf{elim}_B : (a : A) \to (b : B\ a) \to
G\ a\ b\ (\textsf{elim}_A\ a)$. A simpler notion of eliminator does not require
the second component of the motive to mention the first. In particular, we have
$F : A \to \mType$, $G : (a : A) \to B\ a \to \mType$, and eliminators
$\textsf{elim}_A : (a : A) \to F\ a, \textsf{elim}_B : (a : A) (b : B\ a) \to
G\ a\ b$. Note that this elimination principle---that we call \emph{simple
  elimination} after [cite]---is no longer recursive-recursive, since
$\textsf{elim}_A$ is not mentioned in the signature of $\textsf{elim}_B$.

\subsection{Reducing the setoid universe}

The reduction described in the previous section works generically for an
arbitrary finitary inductive-inductive type, thus giving a systematic way to
reduce finitary inductive-inductive definitions to inductive families. However,
it is not clear whether this method extends to infinitary induction-induction,
of which the setoid universe defined in Section~\ref{?} is an instance.
%
Of course, the absence of a general reduction method does not mean that
particular concrete instances of infinitary induction-induction cannot be
reduced to simpler principles, which is exactly what we hope for our universe
construction.

The obvious challenge in successfully completing this reduction is to avoid the
need for extensionality in the metatheory. In fact, consider the simple
infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
soon as we try to define the general eliminator, since it requires function
extensionality. \todo{Is this true?}

Luckily, in our case we can take advantage of our specific circumstances. First
of all, we can define the well-typing predicates as $\mProp$-valued families,
rather than in $\mType$; without this, we would need function extensionality to
prove that the predicates are propositions (with respect to propositional
equality).

\begin{align*}
  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
\end{align*}
\todo{check that these signatures are correct}

Using $\mProp$ would help only to some extent if we had to define general
recursive-recursive eliminators, since that would also require
extensionality. However, it turns out that for our limited use of this
inductive-inductive universe the \emph{simple} elimination principle is
sufficient. This eliminator can be defined by simple pattern matching, and does
not require extensionality.

@@
